Simple connectome inference from partial correlation statistics in calcium imaging

### 2014-06-19 article.tex
Correlation statistics

As the first plot in Figure~\ref{fig:filtered-signal} illustrates, the raw
fluorescence signal is very noisy due to light scattering artifacts that
ordinarily affects the quality of the recording~\citep{lichtman2011big}.
Accordingly, the first step of our pipeline is to smoothen the signal, using
one the following low-pass filter for filtering out high frequency noise :
\begin{align}
% symmetrical median filter
f_1(x^t_i) &= x^{t-1}_i + x^{t}_i + x^{t+1}_i \label{eq:symetric-median}, \\
% asymmetrical weighted median
f_2(x^t_i) &= 0.4 x^{t-3}_i + 0.6 x^{t-2}_i + 0.8 x^{t-1}_i + x_{t}^i,
\label{eq:weighted-asymetric-median}
\end{align}
where \textcolor{red}{$f_?$} effect is shown in the second plot in the figure.

@@

In those terms, the full signal processing pipeline of our approach is defined
by the composed function $w \cdot h \cdot g \cdot f_1$ (resp $f_2$).

@@

Let us denote the fluorescence calcium indicators of all neuron
as a set of random variables $X = \left\{X_1, \ldots, X_{|V|}\right\}$, which follows
a joint probability distribution $p_\mathcal{X}$, and by
$X^{-i,j}$ the set of all variables except $X_i$ and $X_j$.
Two random variables $X_i$ and $X_j$ are said to be conditionally independent
to a (possibly empty) set of random variables $Z$, denoted by $X_i \perp X_j | Z$,
if the joint conditional probability distribution factorizes
$p_{\mathcal{X}_i, \mathcal{X}_j|\mathcal{Z}} = p_{\mathcal{X}_i|\mathcal{Z}}
p_{\mathcal{X}_j|\mathcal{Z}}$.  The inference of the undirected network
can be formulated as inferring conditional dependences
$X_i \not\perp X_j | X^{-i,j}$ between all pairs $i$ and $j$ of fluorescence
signals.

Since all neurons are assumed to form a connected graph, we
have by construction $X_i \not\perp X_j$. However, two random variables $A$ and $B$ might be
dependent taken alone ($A \not\perp B$), but not once we observe the set  $C$
of all other variables ($A \perp B | C$). The conditioning over all other
random variables $X^{-i,j}$ is thus necessary to avoid predicting an edge when
two neurons are not directly connected.

If the random variables $X$ follows a joint Gaussian distribution
$p_\mathcal{X} \sim \mathcal{N}(\mu; \Sigma)$ of mean vector $\mu$ and
covariance matrix $\Sigma$, we have the conditional independence
$X_i \perp X_j | X^{-i,j}$ if the partial correlation coefficient
\[
\rho_{X_i, X_j | X^{-i,j}}
= \frac{\Sigma^{-1}_{ij}}{\sqrt{\Sigma^{-1}_{ii} \Sigma^{-1}_{jj}}}
\]
is equal to zero \citep{koller2009probabilistic}. The partial correlation gives
an estimate of the degree of association between all
pairs of neuron $i$ and $j$.
Similarly, we also have $X_i \perp X_j$ if the Pearson correlation
coefficient
\[
\rho_{X_i,X_j} = \frac{\Sigma_{ij}}{\sqrt{\Sigma_{ii}
\Sigma_{jj}}}
\]
is equal to zero.

% Speak about why averaging
The quality of the signal filtering and network inference depends on
good hyper-parameter selection, which are themselves dependent on the properties
of the targeted network and measurement conditions, e.g. signal to noise ratio.
In order to improve robustness, a weighted average of partial correlation
matrices was estimated from the time series by varying the hard-threshold value
$\tau$ and the design of the low-pass filter (mentionned in section
\ref{sec:filter} and appendix \ref{app:optimized}).

To filter even more noise before computing the partial correlation, we kept
$80\%$ of the eigen vectors which have the largest eigen
value (\cite{tipping1999mixtures}). In the context of the challenge, this
assumption was deduced from ground truth of provided networks.

@@

@@

\subsection*{Partial correlation improves over state-of-the-art methods}

Figure \ref{fig:curves} shows the ROC curve and precision-recall curve
for correlation, GTE and partial correlation on the \textit{normal-2}
dataset. The partial correlation outperforms other state-of-the-art methods
by at least factor of 2 in AUPRC reaching $0.347$ and an increased of $0.05$ in
AUROC.

Two reasons might explain such overwhelming difference: (i) among
those methods only the partial correlation makes the distinction between
direct and indirect connections, (ii) the simulated network from the
\citep{stetter2012model} generator could be reduced to a linear model after
pre-processing. Those results is obviously generalized over the other
datasets see Appendix \textcolor{red}{REF}.

As it can also be seen on Figure \ref{fig:curves}, the GENIE3 performance
increases with deeper trees in addition to be computionnally feasible. The
reason is twofold. First, the deeper the tree, the larger is the conditioning.
As it was observed in the comparison between correlation and partial
correlation, it screens indirect links in the infered network. Second, the
number of small trees need to be much higher than deep trees in order to
observe - and evaluate - all pairs.  However, as suggested by
\cite{louppe2014understanding}, limiting the maximal depth of trees may help
to reduce the misestimation bias in variable importance comuting, i.e.
infering the strengh of a link. Therefore a trade-off of the maximal depth may give the best of both worlds.

% Opt: obviously when parameters are tuned/optimized, it's slightly better -->
% see appendix for a full description of the method

\begin{figure}[bth]
\centering
\includegraphics[width=\linewidth]{images/curves}
\caption{Method precision-recall and roc curves. \textcolor{red}{GENI3?}
         $\tau=0.11$}
\label{fig:curves}
\end{figure}

@@

\textcolor{red}{The whole sub-section should be moved to the appendix and
                replace by one or two sentences.}

@@

It also appears that the averaging over the low-pass filter designs and the
threshold values increases significantly the quality of the inferred network
according to both metrics.

% Explaining the method (thèse de Gilles)
% Avg.: Averaging over the threholds and L-P filters improves a lot both metrics.
% Averaging

\begin{table}[tbh]
\centering
\small
\begin{tabular}{@{}l *{8}{c}@{}}
\hline
  & \multicolumn{4}{c}{AUROC} & \multicolumn{4}{c}{AUPRC} \\
\hline
Methods $\backslash$ normal- & 1 & 2 & 3 & 4 & 1 & 2 & 3 & 4 \\
No  filtering       & 0.777 & 0.767 & 0.772 & 0.774 & 0.070 & 0.064 & 0.068 & 0.072\\
P                   & 0.925 & 0.925 & 0.924 & 0.923 & 0.322 & 0.324 & 0.323 & 0.312\\
P + W               & 0.931 & 0.930 & 0.928 & 0.927 & 0.333 & 0.334 & 0.327 & 0.313\\
P + W + PCA         & 0.932 & 0.931 & 0.930 & 0.928 & 0.364 & 0.366 & 0.359 & 0.344\\
Averaging           & 0.943 & 0.942 & 0.942 & 0.939 & 0.403 & 0.404 & 0.398 & 0.388\\
%Averaging + Directivity & 0.944 & 0.943 & 0.942 & 0.940 & 0.404 & 0.405 & 0.399 & 0.389\\
\end{tabular}
\caption{TODO \textcolor{red}{version simple!!}}
\label{tab:tab3}
\end{table}

\subsection*{Area under the ROC curve is a poor metric for network inference}

The best method only infers the \textbf{undirected} network by giving the
same connectivity score to both edges $(i,j)$ and $(j,i)$. This is quite
surprising considering the problem statement.

If the undirected network is perfectly inferred, it would already yield
high AUROC score for instance $\approx{}0.996$ on \textit{normal-1}. Trying
to infer edge directivity is likely to be counter-productive or yield
small gain.
By contrast, the same experiment with AUPRC would yield
to a score of $\approx 0.789$ on \textit{normal-1}. Thus inferring
correctly directivity would be more rewarding in AUPRC term.
Interestingly, methods that takes into account directivity, namely GTE and
GENIE3, improve their AUROC and AUPRC by giving the same way to directed edges.
Trying to detect the link direction is more penalizing than rewarding.

The area under the ROC curve is not consistent metric
with respect to the problem statement. We advice to avoid the AUROC
for the assessment of network inference method. An alternative metric would
be the area under the precision-recall curve (AUPRC). Those
conclusions are in line with \cite{schrynemackers2013protocols}.

@@

\section{Extended results}

>>>>

As the first plot in Figure~\ref{fig:filtered-signal} illustrates, the raw
fluoresence signal is very noisy due to light scattering artifacts that
ordinarily affects the quality of the recording~\citep{lichtman2011big}.
Accordingly, the first step of our pipeline is to smoothen the signal, using a
low-pass filter for filtering out high frequency noise and resulting in the
second plot in the figure:
\begin{align}
% symmetrical median filter
f_1(x^t_i) &= x^{t-1}_i + x^{t}_i + x^{t+1}_i \label{eq:symetric-median}, \\
% asymmetrical weighted median
f_2(x^t_i) &= 0.4 x^{t-3}_i + 0.6 x^{t-2}_i + 0.8 x^{t-1}_i + x_{t}^i,
\label{eq:weighted-asymetric-median}
\end{align}
where $f_1$ is the filter proposed in the simplified variant of our approach
and $f_2$ is the filter actually used within our winning algorithm.

@@

In those terms, the full signal processing pipeline of our (simplified)
approach is defined by the composed function $w \cdot h \cdot g \cdot f_1$
(resp. $f_2$ for the winning method).

@@

Let us model the fluorescence indicators of all $p$ neurons as a set of random
variables $X = \{X_1, \dots, X_p\}$ following a joint probability distribution
$P_X$. In the statistical framework, two random variables $X_i$ and $X_j$ are
said to be conditionally independent to a (possibly empty) set of controlling
random variables $Z$, denoted $X_i \perp X_j | Z$, if the joint conditional
probability distribution $P_{X_i, X_j | Z}$ factorizes into $P_{X_i|Z}
P_{X_j|Z}$. In particular, when $X_i$ and $X_j$ are conditionally independent
with respect to all the other variables, denoted $X_i \perp X_j | X^{-i,j}$, then the
observation of $X_i$ given the knowledge of $X^{-i,j}$ provides no information
about $X_j$, and vice-versa. From a graphical point of view and assuming no
hidden variable, this indicates that $X_i$ and $X_j$ do not directly interact
(i.e., $(i,j) \notin E$). By opposition, if $X_i$ and $X_j$ are conditionally
dependent with respect to all other variables, denoted $X_i \not\perp X_j |
X^{-i,j}$, then the observation of one of the two variables provides some
information about the other, therefore highlighting a direct interaction
between the two random variables (i.e., $(i, j) \in E$).

Under that assumption that random variables follow a joint Gaussian
distribution $\mathcal{N}(\mu; \Sigma)$ of mean vector $\mu$ and covariance
matrix $\Sigma$, $X_i$ and $X_j$ are conditionnally independent with respect to
the other variables $X^{-i,j}$ if their \textit{partial correlation} statistic
is null~\citep{koller2009probabilistic}, i.e., if
\begin{equation}
\rho_{X_i, X_j | X^{-i,j}} = -\frac{\Sigma^{-1}_{ij}}{\sqrt{\Sigma^{-1}_{ii} \Sigma^{-1}_{jj}}} = 0.
\end{equation}
Conversely, $X_i$ and $X_j$ are conditionally dependent if $\rho_{X_i, X_j | X^{-i,j}} \neq 0$,
where $\rho_{X_i, X_j | X^{-i,j}}$ can now be interpreted as the degree of association
between the two variables (i.e., the larger the value, the stronger the interaction).
Together, the partial correlation statistics correspond to the inverse
$\Sigma^{-1}$ of the covariance matrix, also known as the concentration of
precision matrix. Computationnally, they can be estimated efficiently in
$O(M^3)$ (instead of $O(p^3)$) from the $M$ first principal components
extracted through PCA~\citep{bishop2006pattern} on the processed data. In our experiments, $M$ is set
to $0.8 p$, thereby allowing us to further filter noise in the data.

To highlight the benefits of measuring the partial dependence conditional to
all other variables, we compare in Section~\ref{sec:results} with
\textit{Pearson correlation} statistics, which measure instead the linear dependence
between variables, independently of the others. Accordingly, $X_i$ and $X_j$ are
independent if their Pearson correlation statistic is zero, that is if
\begin{equation}
\rho_{X_i,X_j} = \frac{\Sigma_{ij}}{\sqrt{\Sigma_{ii} \Sigma_{jj}}} = 0.
\end{equation}

Finally, let us remark that performance shows to be quite sensitive the choice
of hyper-parameters in our signal processing pipeline (e.g., $f_1$ or $f_2$ or
the value of the threshold $\tau$). To further improve robustness, the last
step of our method therefore consists in averaging correlation statistics over
various values of the parameters, thereby reducing the variance of our
predictions. Further details are provided in Appendix~\ref{app:optimized}.

@@

%\textcolor{red}{SEUIL 0.11 for tabulars!!!}

% In this section, we will first describe datasets and methods used for the
% comparison. Then, on Table \ref{tab:tab3}, our approach is decomposed to show
% the contribution of each pre-processing steps. Afterwards, we will perform a comparison
% between several methods on Figure \ref{fig:curves} and Table \ref{tab:tab1}. Note that two
% metrics are used to evaluate the performance and a single low-pass filter is
% used (Eq. \ref{eq:symetric-median}) and a threshold, i.e. $\tau = 0.11$. The end
% of this section is devoted to a discussion about the choice of metric and its
% relevance for the challenge.

% \subsection{Assessment methodology}

@@

\subsection*{Comparison with state-of-the art methods}

Figure \ref{fig:curves} shows the ROC curve and precision-recall curve
for correlation, GTE and partial correlation on the \textcolor{red}{normal-?}
dataset. The partial correlation outperforms other state-of-the-art methods
by at least factor of 2 in AUPRC and an increased of $5\%$ in AUROC. There
are two reasons that explains such overwhelming difference: (i) among
those methods only the partial correlation make the distinction between
direct and indirect connection, (ii) the simulated network from the
\citep{stetter2012model} generator could be reduced to a linear model after
pre-processing.

\begin{figure}[bth]
\centering
\includegraphics[width=0.6\linewidth]{images/curves}
\caption{Method precision-recall and roc curves. \textcolor{red}{GENI3?}}
\label{fig:curves}
\end{figure}

Observing Table \ref{tab:tab1}, it appears that best methods only infer the
\textbf{undirected} network by giving the same weights to both edges $(i,j)$
and $(j,i)$. This is quite surprising considering the goal of the challenge,
namely predicted directed connections between neurons. Besides, if the true
network undirected, i.e. for all links $(i,j)$, back links are added, is the
infered network then the score obtained according to the ROC AUC metric is
very high ($\approx 0.9956$ on \textit{normal-1}). Therefore, an undirected
infered network seems sufficient to maximize the ROC AUC score without trying
to identify the link direction. Indeed, predicting perfectly directivity only
yields a small gain ($\approx 0.005$ on \textit{normal-1}) which is is not
very rewarding. In contrast, the same experiment with the precision recall
metric rewards less the undirected network ($\approx 0.789$ on
\textit{normal-1}) and the gain margin in predicting the link directivity is
large enough to be interesting ($\approx 0.211$ on \textit{normal-1}). For
these reason, it seems that the metric used in the contest, i.e. the ROC AUC,
does not seem consistent with the goal of predicting directed connections
between neurons. Another metric, such as the AUC PRC, is maybe more
appropriate to assess the quality of the \textbf{directed} infered network.

Interestingly, an easy way to improve results for asymmetric methods (see
Tab.\ref{tab:tab2}), namely GTE and GENIE3, is to make undirected the infered
network. It proves once again that trying to detect the link direction is more
penalizing than rewarding.

% TODO
%   - discussions about the results
%   - discussions about the metrics (Gilles' text)

% Dummy classifier?

% GTE : Explaning the method (default parameters!)

% Explaining the method (thèse de Gilles)

% Directivity : ?

% Opt: obviously when parameters are tuned/optimized, it's slightly better -->
% see appendix for a full description of the method
% Avg.: Averaging over the threholds and L-P filters improves a lot both metrics.

% Partial correlation : Obvious improvements (both metric) because of the conditioning
% Correlation : Already give good results

% Averaging

\textcolor{red}{TODO: depth discussion }

% Discussion about the metrics

@@

@@

It also appears that the averaging over the low-pass filter designs and the
threshold values increases significantly the quality of the infered network
according to both metrics.

\begin{table}[tbh]
\centering
\begin{tabular}{@{}l *{8}{c}@{}}
\hline
  & \multicolumn{4}{c}{AUROC} & \multicolumn{4}{c}{AUPRC} \\
\hline
Methods $\backslash$ normal- & 1 & 2 & 3 & 4 & 1 & 2 & 3 & 4 \\
No  filtering       & 0.777 & 0.767 & 0.772 & 0.774 & 0.070 & 0.064 & 0.068 & 0.072\\
P                   & 0.925 & 0.925 & 0.924 & 0.923 & 0.322 & 0.324 & 0.323 & 0.312\\
P + W               & 0.931 & 0.930 & 0.928 & 0.927 & 0.333 & 0.334 & 0.327 & 0.313\\
P + W + PCA         & 0.932 & 0.931 & 0.930 & 0.928 & 0.364 & 0.366 & 0.359 & 0.344\\
Averaging           & 0.943 & 0.942 & 0.942 & 0.939 & 0.403 & 0.404 & 0.398 & 0.388\\
%Averaging + Directivity & 0.944 & 0.943 & 0.942 & 0.940 & 0.404 & 0.405 & 0.399 & 0.389\\
\end{tabular}
\caption{TODO \textcolor{red}{version simple!!}}
\label{tab:tab3}
\end{table}

@@
